Stuff

### 2014-01-23 paper/Background.tex
added more explanations
typo errors
reverted to complete proof

In order to obtain the execution semantics of a multi-threaded program on a
multi-processor system, one has to understand the memory consistency model of the
underlying system. This is necessary because, as opposed to a direct access to
a single-ported memory, the processors a) have store buffers which reorders the
sequence of memory requests from the processor, for faster access (say by bypassing
a store-value to a subsequent load request), and b) communicate with the memory
through several levels of caches, which cache the data for faster access. The
memory consistency model describes the semantic behavior of such requests and
the implementation of the system ensures that the cache hierarchy and the store
buffers do not violate the specified semantics.

Memory consistency models can be partitioned into two sub-parts
\cite{Arvind-memory-model}: a) the store-atomicity property for each
location/address and b) the reordering restrictions between requests to
different locations within each processor.  The store-atomicity property per address
can be informally explained as follows: each location behaves like an atomic
register in which no two writes (stores) to that location can happen
simultaneously, and any read of that location (load) returns the previously
written value in that location. Most memory consistency models require that
the memory subsystem obeys the store-atomicity property, which is a global
property for the whole multi-processor system. Unlike the store-atomicity property,
the reordering restrictions dictated by the memory consistency model pertain to
reordering requests within the same processor, but across multiple addresses. For
instance, it dictates whether a store request to an address $x$ can be issued
to the memory sub-system before a load request to a different address $y$,
which was issued by the same processor earlier (\ie load $y$ happened before store
$x$ is program order, but the order of issue to memory sub-system is reversed).

@@

A request issued by a processor to the memory sub-system will contain the
address corresponding to the request, and a type indicating whether the request
is a load or a store. In case of a store request, it will normally contain the
data to be stored. We augment the request message with two special fields: a
\mylabel which uniquely identifies each request, and an \myindex which
is a natural number denoting the position of a request amongst those emanating
from a particular processor. We omit the data field because it can be subsumed
by the unique label (the data and the unique label together forms another
unique label).

A response issued by the memory sub-system to the processor will contain a
label to identify the request corresponding to the response and the processor
to which the response is sent to. In the case of a load request, the memory
sub-system returns \stl,  the label of the last store to that address. In
case there has been no store to that address, a special value called Initial is
returned. The memory sub-system also returns a natural number, \mytime
which represents the position at which this request has been handled globally
among all the requests issued from all the processors. Both the store label (or
Initial) and the time are needed just for the definition of the store-atomicity
property -- a real system will simply return the data corresponding to the load
request's address.

\begin{figure}
\centering
\begin{tabular}{|p{.02\textwidth}lp{.9\textwidth}|}
\hline
\multicolumn{3}{|c|}{\Request}\\
\hline
&\proc:& Processor issuing the request\\
&\mylabel:& A unique label to identify the request\\
&\loc:& Address associated with the request\\
&\desc:& Request type (\Ld{} or \St)\\
&\myindex:& A unique natural number indicating the local order of requests\\
\hline
\hline
\multicolumn{3}{|c|}{\Response}\\
\hline
&\proc:& Processor which receives the response\\
&\mylabel:& Label identifying the request corresponding to this response\\
&\stl:& The label of the last store to that address (or the special value,
       Initial) in case of a load response\\
&\mytime:& A natural number indicating the global order of (processing the) requests\\
\hline
\end{tabular}
\caption{Data types for request and response}
\mylabel{req-resp}
\end{figure}

The uniqueness of request labels (globally) and request indices (per processor) can be defined as follows:
\begin{defn}
\begin{multline*}
\textit{uniqReqLabels} := \forall q_1, q_2\in\Request, q_1.\mylabel = q_2.\mylabel @-> q_1 = q_2\\
\end{multline*}
\end{defn}
\begin{defn}
\begin{multline*}
\textit{uniqIndices} := \forall q_1, q_2\in\Request,
     q_1.\proc = q_2.\proc @-> q_1.\myindex = q_2.\myindex \\ @-> q_1 = q_2
\end{multline*}
\end{defn}

@@

In addition, the global ordering of responses to the same processor should not
violate the local ordering of the corresponding requests.
\begin{defn}
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response) (q_1, q_2 \in \Request), \\ r_1.\mylabel
= q_1.\mylabel @-> r_2.\mylabel = q_2.\mylabel @-> \\ q_1.\proc = q_2.\proc @->
q_1.\loc = q_2.\loc @-> \\ q_1.\myindex < q_2.\myindex @-> r_1.\mytime > r_2.\mytime
  @-> \False
\end{multline*}
\end{defn}

@@

\begin{defn}
\begin{multline*}
\textit{initialValue}:= r.\stl = \Initial \wedge \\
\forall (r' \in \Response) (q' \in \Request),
r'.\mylabel = q'.\mylabel @-> \\ 0 \le r'.\mytime < r.\mytime @->
\neg (q.\loc = q'.\loc \wedge q'.\desc = \St)
\end{multline*}
\end{defn}

\begin{defn}
\begin{multline*}
\textit{lastEarlierValue} :=
\exists m, r.\stl = \Store m \wedge \\
\exists (r_m \in \Response) (q_m \in \Request),
r_m.\mylabel = m \wedge q_m.\mylabel = m \wedge \\
r_m.\mytime < r.\mytime \wedge q_m.\loc = q.\loc \wedge q_m.\desc = \St \wedge  \\
\forall (r' \in \Response) (q' \in \Request),
r'.\mylabel = q'.\mylabel @-> \\ r_m.\mytime < r'.\mytime < r.\mytime ->
\neg (q.\loc = q'.\loc \wedge q'.\desc = \St)
\end{multline*}
\end{defn}

>>>>

In order to obtain the execution semantics of a multi-threaded program on a
multi-processor system, one has to understand the memory consistency model of the
underlying system. This is necessary because, as opposed to a direct access to
a single-ported memory, the processors a) have store buffers which reorders the
sequence of memory requests from the processor, for faster access (say by bypassing
a store-value to a subsequent load request), and b) communicate with the memory
through several levels of data caches for faster access. The
memory consistency model describes the semantic behavior of such requests and
the system ensures that the cache hierarchy and the store
buffers do not violate the specified semantics.

Memory consistency models can be partitioned into two sub-parts
\cite{Arvind-memory-model}: a) the ``global'' store-atomicity property for each
location/address and b) the ``local'' reordering restrictions between requests
to different locations within each processor. The store-atomicity property per
address can be informally explained as follows: each location behaves like an
atomic register in which no two writes (stores) to that location can happen
simultaneously, and any read of that location (load) returns the previously
written value in that location. Most memory consistency models require that the
memory subsystem obeys the store-atomicity property. The reordering
restrictions dictated by the memory consistency model pertain to reordering
requests within the same processor (unlike the store-atomicty property which is
global across the entire system), but across multiple addresses. For example,
it dictates whether a store request to an address $x$ can be issued to the
memory sub-system before a load request to a different address $y$, which was
generated by the same processor earlier (\ie load $y$ happened before store $x$
is program order, but the order of issue to memory sub-system is reversed).

@@

Figure \ref{req-resp} shows the data-types corresponding to a request issued
from the processor to the memory subsystem and a response issued from the
memory sub-system to the processor. A request contains the
address corresponding to the request, and a type indicating whether the request
is a load or a store. In case of a store request, a real system will contain the
data to be stored. We augment the request message with two special fields: a
\mylabel which uniquely identifies each request, and an \myindex which
is a natural number denoting the position of a request amongst those emanating
from a particular processor. We omit the data field because it can be subsumed
by the unique label (the data and the unique label together forms another
unique label).

A response issued contains a label to identify the request corresponding to the
response and the processor to which the response is sent to. In the case of a
load request, the memory sub-system returns \stl, the label of the last store
request to that address. In case there has been no store to that address, a
special value called Initial is returned. The memory sub-system also returns a
natural number, \mytime which represents the position at which this request has
been handled globally among all the requests issued from all the processors.
Both the store label (or Initial) and the time are needed just for the
definition of the store-atomicity property -- a real system will simply return
the data corresponding to the load request's address.

\begin{figure}
\centering
\begin{tabular}{|p{.02\textwidth}lp{.9\textwidth}|}
\hline
\multicolumn{3}{|c|}{\Request}\\
\hline
&\proc:& Processor issuing the request\\
&\mylabel:& A unique label to identify the request\\
&\loc:& Address associated with the request\\
&\desc:& Request type (\Ld{} or \St)\\
&\myindex:& A unique natural number indicating the local order of requests\\
\hline
\hline
\multicolumn{3}{|c|}{\Response}\\
\hline
&\proc:& Processor which receives the response\\
&\mylabel:& Label identifying the request corresponding to this response\\
&\stl:& The label of the last store to that address (or the special value,
       Initial) in case of a load response\\
&\mytime:& A natural number indicating the global order of (processing the) requests\\
\hline
\end{tabular}
\caption{Data types for request and response}
\label{req-resp}
\end{figure}

The \Request set in the following contains only those requests processed by the
memory subsystem and the \Response set contains only the responses generated by
the memory subsystem.

The uniqueness of request labels (globally), which is needed for defining store
atomicity, can be defined as follows:
\begin{defn}
\begin{multline*}
\textit{uniqReqLabels} := \forall q_1, q_2\in\Request, q_1.\mylabel = q_2.\mylabel @-> q_1 = q_2\\
\end{multline*}
\end{defn}

Requests from the same processor to the memory subsystem are sent in the order
of their \myindex field. Thus, is request $r$ is sent to a processor $c$, then
the next request that can be sent to the processor obeys the following
property.
\begin{defn}
\begin{multline*}
\textit{nextReq} r c := 
\end{multline*}
\end{defn}

@@

In addition, the global ordering of responses to the same processor should not
violate the local ordering of the corresponding requests.
\begin{defn}
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response) (q_1, q_2 \in \Request), \\ r_1.\mylabel
= q_1.\mylabel @-> r_2.\mylabel = q_2.\mylabel @-> \\ q_1.\proc = q_2.\proc @->
q_1.\loc = q_2.\loc @-> \\ q_1.\myindex < q_2.\myindex @-> r_1.\mytime \le r_2.\mytime
\end{multline*}
\end{defn}

@@

\begin{defn}
\begin{multline*}
\textit{initialValue}:= \forall (r \in \Request) (q \in \Response), r.\mylabel = q.\mylabel @->
q.\desc = \Ld @-> \\(r.\stl = \Initial \wedge
\forall (r' \in \Response) (q' \in \Request),
r'.\mylabel = q'.\mylabel @-> \\ 0 \le r'.\mytime < r.\mytime @->
\neg (q.\loc = q'.\loc \wedge q'.\desc = \St))
\end{multline*}
\end{defn}

\begin{defn}
\begin{multline*}
\textit{lastEarlierValue} := \forall (r \in \Request) (q \in \Response), r.\mylabel = q.\mylabel @-> \\
q.\desc = \Ld @-> 
(\exists m, r.\stl = \Store m \wedge \\
\exists (r_m \in \Response) (q_m \in \Request), 
r_m.\mylabel = m \wedge q_m.\mylabel = m \wedge \\
r_m.\mytime < r.\mytime \wedge q_m.\loc = q.\loc \wedge q_m.\desc = \St \wedge  \\
\forall (r' \in \Response) (q' \in \Request),
r'.\mylabel = q'.\mylabel @-> \\ r_m.\mytime < r'.\mytime < r.\mytime ->
\neg (q.\loc = q'.\loc \wedge q'.\desc = \St))
\end{multline*}
\end{defn}
